Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-02-01 ATACCC.tex
include false negative estimates
Use RT-PCR (not PCR)

\Cref{E-inc-prev} showed that the duration of PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
The aim of this chapter is to estimate the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
I model a biomarker, the log viral load, over the course of an infection.
From this model, I derive the distribution of the duration of PCR positivity.

First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before modifying its observation model to improve the inference (\cref{ATACCC:sec:observation-modification}).
This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of PCR positivity (\cref{ATACCC:sec:duration}).
I will refine this estimate in future chapters before its use in estimating transmission.

@@

The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
The viral load model produces a deterministic trajectory for the viral load of an individual, a function of the model parameters.
The observation model then describes how the PCR testing process produces noisy observations of this viral load.
The observation model includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
% The limit of detection means that viral loads below a certain threshold are not detected by the PCR test.
% They can be represented as a censoring of the viral load at the limit of detection.

@@

PCR tests are noisy observations of the viral load.
A positive PCR test has a Ct value.
Ct values are linearly related to viral load (see \cref{E-biology-data:sec:PCR})
A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
Positive PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.

@@

The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of PCR positivity.
So far, only the viral load trajectories have been considered; \textcite{hakkiOnset} does not consider this duration either.
The time that individual $i$ are first and last positive are the two values of $\tau$ such that $\log v_i(\tau) = v_\text{lod}$.
Denote these times as $\tau_{0,i}$ and $\tau_{1,i}$.
Hence, their duration of PCR positivity, in continuous time, is $d_i = \tau_{1,i} - \tau_{0,i}$.

@@

The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits}).
Almost all test results are captured in the 95\% credible intervals.
Individual 55 has some fitting issues, explored further below.
Individual 57 appears to have far noisier measurements than the other individuals, suggesting that the observation noise might vary between individuals.
Including a random effect on $\sigma_v$ could capture this affect.
However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
\begin{figure}
  \thisfloatpagestyle{empty}
  \vspace{-3cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
  \captionsetup{width=0.8\paperwidth}
  \captionof{figure}[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for each individual, overlaid with their observed test results (dots). Posterior estimates are conditional on the result not being a false negative. Negative PCR tests are plotted at the limit of detection, \ie a log viral load of 0. Each plot is labelled with the individual's position in the dataset, an arbitrary value. \label{ATACCC:fig:goodness-of-fits}}
\end{figure}

@@

Clear differences emerge between the vaccinated and unvaccinated groups (see \cref{ATACCC:fig:median-trajectories} and \cref{ATACCC:table:population-estimates}).
The rates of viral load increase at similar rates in the two groups, the vaccinated group have higher peak viral loads and clear the virus faster.
This leads to a shorter duration of PCR positivity in the vaccinated group.
\begin{figure}
  \centering \includegraphics{ATACCC/mean_trajectories}
  \caption[Median viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation. These assume $\tau_{\max,i}=0$, however modifying this simply translates the trajectories left or right. \label{ATACCC:fig:median-trajectories}}
\end{figure}
\begin{table}
  \makebox[\linewidth]{
  \begin{tabular}{lcc}
    Parameter & Unvaccinated population & Vaccinated population \\
    Max log viral load, $\mu_{v_\text{max},g}$ & 14.9 (14.2--15.7) & 16.0 (15.2--16.8) \\
    Rate of log viral load increase, $\exp(\mu_{a,g})$ & 5.07 (3.70--7.06) & 1.25 (1.18--1.32) \\
    Rate of log viral load decrease, $\exp(\mu_{b,g})$ & 1.33 (1.09--1.63) & 1.89 (1.50--2.36) \\
  \end{tabular}
  }
  \caption{Population-level estimates of the viral load parameters (posterior medians and 95\% CrI). The vaccinated population have higher peak viral loads but clear the virus faster. \label{ATACCC:table:population-estimates}}
\end{table}

The negative rate, $\rho$, is estimated to be quite low, 5.3\% (95\% CrI: 3.7--7.3\%).

The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
In \cref{ATACCC:sec:results}, the two models gave very similar results.
Therefore, I show results only from the modified model (which had better convergence).
\begin{figure}
  \centering \includegraphics{ATACCC/duration}
  \caption[Duration of PCR positivity.]{Estimated distribution (survival function) of duration of PCR positivity for the two populations. The vaccinated population have shorter durations of PCR positivity.  \label{ATACCC:fig:duration}}
\end{figure}

The vaccinated population have shorter durations of PCR positivity than the unvaccinated population (see \cref{ATACCC:fig:duration}).
The median duration of PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.

@@

This chapter has estimated the distribution of PCR positive duration based on the viral load data from the ATACCC study.

>>>>

\Cref{E-inc-prev} showed that the duration of RT-PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
The aim of this chapter is to estimate the duration of RT-PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
I model a biomarker, the log viral load, over the course of an infection.
From this model, I derive the distribution of the duration of RT-PCR positivity.

First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before modifying its observation model to improve the inference (\cref{ATACCC:sec:observation-modification}).
This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of RT-PCR positivity (\cref{ATACCC:sec:duration}).
I will refine this estimate in future chapters before its use in estimating transmission.

@@

The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
The viral load model produces a deterministic trajectory for the viral load of an individual, a function of the model parameters.
The observation model then describes how the RT-PCR testing process produces noisy observations of this viral load.
The observation model includes the possibility of false positive and negatives, and the limit of detection of the RT-PCR test.
% The limit of detection means that viral loads below a certain threshold are not detected by the RT-PCR test.
% They can be represented as a censoring of the viral load at the limit of detection.

@@

RT-PCR tests are noisy observations of the viral load.
A positive RT-PCR test has a Ct value.
Ct values are linearly related to viral load (see \cref{E-biology-data:sec:PCR})
A negative RT-PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
Positive RT-PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.

@@

The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of RT-PCR positivity.
So far, only the viral load trajectories have been considered; \textcite{hakkiOnset} does not consider this duration either.
The time that individual $i$ are first and last positive are the two values of $\tau$ such that $\log v_i(\tau) = v_\text{lod}$.
Denote these times as $\tau_{0,i}$ and $\tau_{1,i}$.
Hence, their duration of RT-PCR positivity, in continuous time, is $d_i = \tau_{1,i} - \tau_{0,i}$.

@@

The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits}).
Almost all test results are captured in the 95\% credible intervals.
Individual 55 has some fitting issues, explored further below.
Individual 57 appears to have far noisier measurements than the other individuals, suggesting that the observation noise might vary between individuals.
Including a random effect on $\sigma_v$ could capture this affect.
However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
\begin{figure}
  \thisfloatpagestyle{empty}
  \vspace{-3cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
  \captionsetup{width=0.8\paperwidth}
  \captionof{figure}[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for each individual, overlaid with their observed test results (dots). Posterior estimates are conditional on the result not being a false negative. Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0. Each plot is labelled with the individual's position in the dataset, an arbitrary value. \label{ATACCC:fig:goodness-of-fits}}
\end{figure}

@@

Clear differences emerge between the vaccinated and unvaccinated groups (see \cref{ATACCC:fig:median-trajectories} and \cref{ATACCC:table:population-estimates}).
The rates of viral load increase at similar rates in the two groups, the vaccinated group have higher peak viral loads and clear the virus faster.
This leads to a shorter duration of RT-PCR positivity in the vaccinated group.
\begin{figure}
  \centering \includegraphics{ATACCC/mean_trajectories}
  \caption[Median viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation. These assume $\tau_{\max,i}=0$, however modifying this simply translates the trajectories left or right. \label{ATACCC:fig:median-trajectories}}
\end{figure}
\begin{table}
  \makebox[\linewidth]{
  \begin{tabular}{lcc}
    Parameter & Unvaccinated population & Vaccinated population \\
    Max log viral load, $\mu_{v_\text{max},g}$ & 14.9 (14.2--15.7) & 16.0 (15.2--16.8) \\
    Rate of log viral load increase, $\exp(\mu_{a,g})$ & 5.07 (3.70--7.06) & 1.25 (1.18--1.32) \\
    Rate of log viral load decrease, $\exp(\mu_{b,g})$ & 1.33 (1.09--1.63) & 1.89 (1.50--2.36) \\
  \end{tabular}
  }
  \caption{Population-level estimates of the viral load parameters (posterior medians and 95\% CrI). The vaccinated population have higher peak viral loads but clear the virus faster. \label{ATACCC:table:population-estimates}}
\end{table}

The false negative rate, $\rho$, is estimated to be quite low, 5.3\% (95\% CrI: 3.7--7.3\%).

The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
In \cref{ATACCC:sec:results}, the two models gave very similar results.
Therefore, I show results only from the modified model (which had better convergence).
\begin{figure}
  \centering \includegraphics{ATACCC/duration}
  \caption[Duration of RT-PCR positivity.]{Estimated distribution (survival function) of duration of RT-PCR positivity for the two populations. The vaccinated population have shorter durations of RT-PCR positivity.  \label{ATACCC:fig:duration}}
\end{figure}

The vaccinated population have shorter durations of RT-PCR positivity than the unvaccinated population (see \cref{ATACCC:fig:duration}).
The median duration of RT-PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.

@@

This chapter has estimated the distribution of RT-PCR positive duration based on the viral load data from the ATACCC study.

### 2024-02-01 cis-imperfect-testing.tex
Use RT-PCR (not PCR)

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see  for more details).
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of 1000, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data is shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

In this chapter, I developed and applied novel methodology to estimate the duration distribution of PCR-positivity using information from two studies: ATACCC and CIS.
Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.

These estimates are the first estimates of the duration of PCR-positivity, including the distribution tails, estimated from general population data.
This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting PCR test results.
\todo{possibly cut some of this text?}
Producing these estimates required applying previous survival analysis frameworks in the novel context of the CIS.
The framework needed further modification to account for the presence of false negatives, done in this chapter.

>>>>

% Discussion and interrogation of the data led me to
I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
There are several strands of evidence supporting this hypothesis.
First, it is well known that RT-PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see  for more details).
This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
\begin{figure}
  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
  \caption[Single positive episodes in CIS simulation]{%
    Proportion of detected episodes which have a single positive episode.
    y-axis is the number of simulations, out of 1000, with the indicated proportion.
    Test sensitivity is $\psens$.
    The CIS data is shown as a vertical line.
  }
  \label{imperf-test:fig:sim-single-pos}
\end{figure}

@@

In this chapter, I developed and applied novel methodology to estimate the duration distribution of RT-PCR-positivity using information from two studies: ATACCC and CIS.
Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.

These estimates are the first estimates of the duration of RT-PCR-positivity, including the distribution tails, estimated from general population data.
This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting RT-PCR test results.
\todo{possibly cut some of this text?}
Producing these estimates required applying previous survival analysis frameworks in the novel context of the CIS.
The framework needed further modification to account for the presence of false negatives, done in this chapter.

### 2024-02-01 incidence-prevalence.tex
Use RT-PCR (not PCR)

This chapter defines incidence and prevalence, and the relationship between these quantities.
In the course of explaining this relationship, it will become clear that the duration of PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.

@@

\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
Here, an individual is defined to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive PCR test.
I refer to these individuals as \emph{prevalent} individuals.
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$.
Therefore, the prevalence is $P_t/\Npop$.

@@

The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation.
The equation is of the same as the equation used for HIV/AIDS backcalculation, which itself is often found in a variety of applications~\autocite[and references therein]{brookmeyerBackcalculation}.
The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{L} \vec{Z}$ where $\matr{L}$ is a lower-triangular matrix.
Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
However, these algorithms can be unstable, meaning that numerical errors are large.
This is a particular issue if $S(1)$ is small.
Since, by definition, $S(1)=1$ here, this is not an issue in this context.
%However, if instead of estimating the incidence of PCR positives, the aim was to estimate the incidence of infections, then $S(1)$ would be small and this would be an issue.
%While this quantity is more epidemiologically relevant, the time from infection to PCR positivity is short and, for now, I will neglect the difference.

@@

Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the objective of this thesis.
This chapter has shown that the key quantity relating these quantities is the distribution of the duration of PCR positivity.
However, no estimates of this quantity in the general population exist.
Therefore, I now turn to estimating the duration of PCR positivity.

>>>>

This chapter defines incidence and prevalence, and the relationship between these quantities.
In the course of explaining this relationship, it will become clear that the duration of RT-PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.

@@

\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
Here, an individual is defined to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive RT-PCR test.
I refer to these individuals as \emph{prevalent} individuals.
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$.
Therefore, the prevalence is $P_t/\Npop$.

@@

The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation.
The equation is of the same as the equation used for HIV/AIDS backcalculation, which itself is often found in a variety of applications~\autocite[and references therein]{brookmeyerBackcalculation}.
The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{L} \vec{Z}$ where $\matr{L}$ is a lower-triangular matrix.
Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
However, these algorithms can be unstable, meaning that numerical errors are large.
This is a particular issue if $S(1)$ is small.
Since, by definition, $S(1)=1$ here, this is not an issue in this context.
%However, if instead of estimating the incidence of RT-PCR positives, the aim was to estimate the incidence of infections, then $S(1)$ would be small and this would be an issue.
%While this quantity is more epidemiologically relevant, the time from infection to RT-PCR positivity is short and, for now, I will neglect the difference.

@@

Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the objective of this thesis.
This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.
However, no estimates of this quantity in the general population exist.
Therefore, I now turn to estimating the duration of RT-PCR positivity.

### 2024-02-01 introduction.tex
Use RT-PCR (not PCR)

Two approaches are commonly used to estimate durations: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
For SARS-CoV-2, both approaches would be based on the results of PCR testing.
Modelling a biomarker would mean modelling the viral load, measured as a Ct value (see \cref{E-ATACCC}).
Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted (see \cref{E-perf-test,E-imperf-test}).
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
Modelling a biomarker uses more information per observation, but requires stronger assumptions.

@@

\begin{itemize}
    \item Estimate of duration of PCR-positivity: of interest for rest of thesis but also more broadly for public health (\eg interpreting PCR test results)
    \item Use in interpreting test results
    \item Incidence estimates from only CIS
    \item Transmission estimates from only CIS
\end{itemize}

>>>>

Two approaches are commonly used to estimate durations: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing.
Modelling a biomarker would mean modelling the viral load, measured as a Ct value (see \cref{E-ATACCC}).
Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted (see \cref{E-perf-test,E-imperf-test}).
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
Modelling a biomarker uses more information per observation, but requires stronger assumptions.

@@

\begin{itemize}
    \item Estimate of duration of RT-PCR-positivity: of interest for rest of thesis but also more broadly for public health (\eg interpreting RT-PCR test results)
    \item Use in interpreting test results
    \item Incidence estimates from only CIS
    \item Transmission estimates from only CIS
\end{itemize}

### 2024-02-01 transmission.tex
Use RT-PCR (not PCR)

In this chapter, I use the preferred estimate of the duration of PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
This estimation is based only on the CIS data (see \cref{transmission:sec:data} for further details).
I will show that this data is sufficient to estimate the parameters of the transmission process, suggesting surveys of this type are sufficient for surveillance of an epidemic.

@@

In MRP, a Bayesian multilevel logistic regression model is first fit to predict PCR positivity from demographic covariates.
Then, the posterior distribution is used to predict the prevalence in each stratum.
Finally, the prevalence is poststratified (as explained later) to the CIS target population and subpopulations of interest.
I modify this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
% I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
% Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.

@@

% Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
% The proportion of PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

@@

I link the modelled proportion of individuals that are PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.

@@

    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
\end{tikzpicture}
}
  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
  \label{SEIR:fig:full-model}
\end{figure}

@@

Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev-young}
\end{figure}
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
    }
    \label{SEIR:fig:prev-old}
\end{figure}

@@

\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
Generally the estimates agree, with the SEIR model having much narrower credible intervals.
The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
However, except in London, the credible intervals still overlap.
This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.;
\begin{figure}
    \centering \includegraphics{transmission/compare-regions}
    \caption[Comparing each models estimate of regional incidence.]{%
        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-regions}
\end{figure}

@@

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}

>>>>

In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
This estimation is based only on the CIS data (see \cref{transmission:sec:data} for further details).
I will show that this data is sufficient to estimate the parameters of the transmission process, suggesting surveys of this type are sufficient for surveillance of an epidemic.

@@

In MRP, a Bayesian multilevel logistic regression model is first fit to predict RT-PCR positivity from demographic covariates.
Then, the posterior distribution is used to predict the prevalence in each stratum.
Finally, the prevalence is poststratified (as explained later) to the CIS target population and subpopulations of interest.
I modify this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
% I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
% Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.

@@

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

@@

I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.

@@

    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
\end{tikzpicture}
}
  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents RT-PCR-positive individuals. Shown for a single stratum.}
  \label{SEIR:fig:full-model}
\end{figure}

@@

Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata RT-PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev-young}
\end{figure}
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
    }
    \label{SEIR:fig:prev-old}
\end{figure}

@@

\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
Generally the estimates agree, with the SEIR model having much narrower credible intervals.
The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
However, except in London, the credible intervals still overlap.
This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.;
\begin{figure}
    \centering \includegraphics{transmission/compare-regions}
    \caption[Comparing each models estimate of regional incidence.]{%
        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-regions}
\end{figure}

@@

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}
